New lower bounds for Schur and weak Schur numbers

### 2021-04-08 article/intro.tex
notations :'(

Let us notice that a sum-free subset is also weakly sum-free, hence justifying the name of \textit{weakly} sum-free
subsets. Given \(m\) and \(n\) two integers, we are interested in partitioning the set of integers from 1 to \(m\) into
\(n\) weakly sum-free subsets.

\begin{notation}
We denote by \([\![1,n]\!]\) the set of integers \(\{1, 2, ..., n\}\).
\end{notation}

Schur proved in \hyperlink{label1}{\textbf{[1]}} that given a number of subsets \(n\), there exists a value of \(m\)
such that there exists no partition of \([\![1,p]\!]\) into \(n\) sum-free subsets for any \(p \geqslant m\). A similar
property holds for weakly sum-free subsets (reference necessaire). These observations lead to the following definitions.

@@

\begin{notation}
For a partition of \([\![1, n]\!]\) in \(k\) subsets, we generally note these subsets \(A_1, ..., A_k\). We also note \(m_i = \text{min}(A_i)\). 
By ordering the subsets, we mean assuming that \(m_1 < ... < m_k\). However, if not specified we do not make this hypothesis since we 
do not always consider partitions in which every subset plays a symmetric role.
\end{notation}

>>>>

Let us notice that a sum-free subset is also weakly sum-free, hence justifying the name of \textit{weakly} sum-free
subsets. Given \(p\) and \(n\) two integers, we are interested in partitioning the set of integers from 1 to \(p\) into
\(n\) weakly sum-free subsets.

\begin{notation}
We denote by \([\![1,p]\!]\) the set of integers \(\{1, 2, ..., p\}\).
\end{notation}

Schur proved in \hyperlink{label1}{\textbf{[1]}} that given a number of subsets \(n\), there exists a value of \(p\)
such that there exists no partition of \([\![1,q]\!]\) into \(n\) sum-free subsets for any \(q \geqslant p\). A similar
property holds for weakly sum-free subsets (reference necessaire). These observations lead to the following definitions.

@@

\begin{notation}
For a partition of \([\![1, p]\!]\) in \(n\) subsets, we generally note these subsets \(A_1, ..., A_n\). We also note \(m_i = \text{min}(A_i)\). 
By ordering the subsets, we mean assuming that \(m_1 < ... < m_n\). However, if not specified we do not make this hypothesis since we 
do not always consider partitions in which every subset plays a symmetric role.
\end{notation}

### 2021-04-08 article/sat.tex
notations :'(

We then encode the existence of a partition of \([\![1,n]\!]\) in \(k\)  weakly sum-free subsets as follows: for every integer
\(i \in [\![1,n]\!]\), take \(k\) variables \(x^{(i)}_{1}, ..., x^{(i)}_{k}\) and for every \(\forall c \in [\![1,k]\!], x^{(i)}_c = 1 \iff i \in A_c\).
The corresponding clauses are:

\begin{itemize}
\item \textbf{sumfree:} \(\forall c \in  [\![1,k]\!], \forall (i, j) \in [\![1,n]\!]^2, (i \neq j ~ \text{and} ~ i + j \leq n) \implies \lnot x^{(i)}_c
\lor  \lnot x^{(i)}_c \lor \lnot x^{(i+j)}_c\)
\item \textbf{union:} \(\forall i \in [\![1,n]\!], x^{(i)}_1 \lor ... \lor x^{(i)}_k\)
\item \textbf{disjoint:} \(\forall i \in [\![1,n]\!],\forall (c_1, c_2) \in  [\![1,k]\!]^2, c_1 \neq c_2 \implies \neg x^{(i)}_{c_1} \lor \neg x^{(i)}_{c_2}\)
\end{itemize}

@@

\begin{computational theorem}
There is no weakly sum-free partition of \([\![1,583]\!]\) in 6 parts such that:
\begin{itemize}
	\item \(m_5 \ge 66\)
	\item \(m_6 \ge 186\)
	\item \([\![210,349]\!] \subset A_6\)
\end{itemize}
\end{computational theorem}

@@

\par
The above theorem shows that the previous lower bound for \(WS(6)\) is optimal in the search space considered by the papers which found it.
Therefore, finding a partition of \([\![1,n]\!]\) in 6 weakly sum-free subsets for some \(n \geq 590\) which does not have a template-like structure
would be extremely interesting since it could give indications on a new search space for improving lower bounds with a computer. More generally,
it questions the search space previously used for finding lower bounds for \(WS(n)\) with a computer. In particular, to our knowledge every paper
that found the lower bound \(WS(5) \geq 196\) used this assumption [reference necessaire]. Therefore one may wonder if this actually a good lower
bound. In the next subsection, we give properties that a partition of \([\![1,197]\!]\) in 5 weakly sum-free subsets has to verify.

\subsection{Weak Schur number five}
As explained in the previous subsection, the search space used for showing that  \(WS(5) \geq 196\) may not contain optimal solution. In this subsection,
we give necessary conditions for a hypothetical partition of \([\![1,197]\!]\) in 5 weakly sum-free subsets using the same type of methods as in the 
previous subsection.

\begin{notation}
Let \(P\) be a predicate over weakly sum-free partitions. We denote by \(WS(n | P)\) the greatest number \(N\) such that there exists a partition of  
\([\![1,N]\!]\) in n weakly sum-free subsets which verifies P.
\end{notation}

>>>>

We then encode the existence of a partition of \([\![1,p]\!]\) in \(k\)  weakly sum-free subsets as follows: for every integer
\(i \in [\![1,p]\!]\), take \(k\) variables \(x^{(i)}_{1}, ..., x^{(i)}_{k}\) and for every \(\forall c \in [\![1,k]\!], x^{(i)}_c = 1 \iff i \in A_c\).
The corresponding clauses are:

\begin{itemize}
\item \textbf{sumfree:} \(\forall c \in  [\![1,k]\!], \forall (i, j) \in [\![1,p]\!]^2, (i \neq j ~ \text{and} ~ i + j \leq n) \implies \lnot x^{(i)}_c
\lor  \lnot x^{(i)}_c \lor \lnot x^{(i+j)}_c\)
\item \textbf{union:} \(\forall i \in [\![1,p]\!], x^{(i)}_1 \lor ... \lor x^{(i)}_k\)
\item \textbf{disjoint:} \(\forall i \in [\![1,p]\!],\forall (c_1, c_2) \in  [\![1,k]\!]^2, c_1 \neq c_2 \implies \neg x^{(i)}_{c_1} \lor \neg x^{(i)}_{c_2}\)
\end{itemize}

@@

\begin{computational theorem}
There is no weakly sum-free partition of \([\![1,583]\!]\) in 6 parts such that:
\begin{itemize}
	\item \(m_5 \geqslant 66\)
	\item \(m_6 \geqslant 186\)
	\item \([\![210,349]\!] \subset A_6\)
\end{itemize}
\end{computational theorem}

@@

\par
The above theorem shows that the previous lower bound for \(WS(6)\) is optimal in the search space considered by the papers which found it.
Therefore, finding a partition of \([\![1,n]\!]\) in 6 weakly sum-free subsets for some \(n \geqslant 590\) which does not have a template-like structure
would be extremely interesting since it could give indications on a new search space for improving lower bounds with a computer. More generally,
it questions the search space previously used for finding lower bounds for \(WS(n)\) with a computer. In particular, to our knowledge every paper
that found the lower bound \(WS(5) \geqslant 196\) used this assumption [reference necessaire]. Therefore one may wonder if this actually a good lower
bound. In the next subsection, we give properties that a partition of \([\![1,197]\!]\) in 5 weakly sum-free subsets has to verify.

\subsection{Weak Schur number five}
As explained in the previous subsection, the search space used for showing that  \(WS(5) \geqslant 196\) may not contain optimal solution. In this subsection,
we give necessary conditions for a hypothetical partition of \([\![1,197]\!]\) in 5 weakly sum-free subsets using the same type of methods as in the 
previous subsection.

\begin{notation}
Let \(P\) be a predicate over weakly sum-free partitions. We denote by \(WS(n | P)\) the greatest number \(p\) such that there exists a partition of  
\([\![1,p]\!]\) in n weakly sum-free subsets which verifies P.
\end{notation}

### 2021-04-08 article/schur.tex
notations :'(

\qquad In this section, we use Rowley's constructions [7] in a Schur context. To improve lower bounds for Ramsay's numbers, Rowley 
introduces partitions verifying some properties which can be extended using a method which generalizes Abbott-Hanson's [8] construction. 
Rolwey named these partitions "templates", and we will keep this name in the entire article. We find suitable templates and use them
 to find new lower bounds for Schur numbers.

\subsection{Definition of SE\(_+\)}

\begin{definition}
	We call SF-template of \(k\) colors and lenght \(n\) a partition of \( [\![1,n]\!]\) into \(k\) sum-free subsets \(A_1, A_2, ..., A_k\) which verify :
	\[
	\forall i \in [\![1, k-1]\!], \forall (x,y) \in A_i^2, x+y > n
	\Longrightarrow x+y-n \notin A_i
	\]
	We note \(SE_+(k)\) the maximal lenght for a SF-template of \(k\) colors. 
\end{definition}

\begin{theorem}
	Let \(n \in [\![2, +\infty]\!]\), we have :
	\[
	2S(n-1)+1 \leqslant SE_+(n) \leqslant S(n)
	\]
\end{theorem}

\begin{remark}
	\(SE_+\) and \(S\) have the same asymptotic growth rate.
\end{remark}

\subsection{Inequalities on \(SE_+\)}

The main result on \(SE_+\) follows. It allows us to improve lower bounds on Schur numbers by computing \(SE_+\).

\begin{theorem}
	Let $(n,k), (r,s) \in (\mathbb{N}^*)^2$. If there exists a \(SF\)-template of $k+1$ colors and lenght \(r\),
	and a partition of $n$ sum-free subsets of $[\![1,s]\!]$ then there exists a partition of $n+k$ sum-free subsets of $[\![1,rs]\!]$.
\end{theorem}

Setting $r = SE_+(k+1)$ and $s = S(n)$ yields the following corollary.

\begin{corollary}
	Let \(n, k \in \mathbb{N}^*\), we have \\
	\[ S(n+k) \geqslant SE_+(k+1)S(n) \]
\end{corollary}

Using a SAT solvers, we are able to exhibit SF-template,
hence providing lower bound on \(SE_+\). Moreover, when using 
a SF-template in the above formula, it is possible to get an additive constant. We seek templates providing "affine"
inequalities : \(aS(n+k) + b \) where \(a\) is a lower bound for \(SE_+(k+1)\). Further details can be found in the \hyperref[SAT]{SAT section}.
Here are the best inequalities on Schur numbers so far :
\[ S(n+1) \geqslant 3S(n) + 1 \]
\[ S(n+2) \geqslant 9S(n) + 4 \]
\[ S(n+3) \geqslant 33S(n) + 6 \]
\[ S(n+4) \geqslant 111S(n) + 43 \]
\[ S(n+5) \geqslant 380S(n) + 148 \]
\[ S(n+6) \geqslant 1140S(n) + 528 \]

The first inequality comes from the original Schur's paper \hyperlink{label1}{\textbf{[1]}}. The second and third ones were found by Rowley (reference nécessaire), 
while the other ones are new. \\

We also have a similar theorem where only \(SE_+\) is involved.

\begin{theorem}
	Let $(n,k), (r,s) \in (\mathbb{N}^*)^2$. If there exists a \(SF\)-template of $k+1$ colors and lenght \(r\),
	and \(SF\)-template of \(n\) color and lenght \(s\), then there exists \(SF\)-template of \((n+k)\) and lenght \(rs\).
\end{theorem}

@@

\begin{corollary}
	Let \(n, k \in \mathbb{N}^*\), we have \\
	\[ SE_+(n+k) \geqslant SE_+(k+1)SE_+(n) \]
\end{corollary}

@@

The previous inequalities give new lower bounds for \(S(n)\) for
\( n \in [\![9,+\infty]\!] \backslash \{13\} \). We compute the lower
bounds for \( n \in [\![8,15]\!] \) using the four different inequalities, please notice that the best values for \( n = 8\) and \(n = 13\) were already obtained thanks to the first one, found by Rowley. The best lower bounds are highlighted.\\
\\
\begin{center}
\begin{tabular}{|*{5}{c|}}
    \hline
	 & 8 & 9 & 10 & 11 \\
	\hline
	\(33S(n-3) + 6 \) & \cellcolor{yellow} 5286 & 17694 & 55446 & 174444\\
	\hline
	\(111S(n-4) + 43 \) & 4927 & \cellcolor{yellow} 17803 & 59539 & 186523\\
	\hline
	\(380S(n-5) + 148 \) & 5088 & 16868 & \cellcolor{yellow} 60948 & \cellcolor{yellow} 203828 \\
	\hline
	\(1140S(n-6) + 528 \) & 5088 & 15348 & 50688 & 182928\\
	\hline
	\hline
	& 12 & 13 & 14 & 15 \\
	\hline
	\(33S(n-3) + 6 \) & 587505 & \cellcolor{yellow} 2011290 & 6726330 & 21072090\\
	\hline
	\(111S(n-4) + 43 \) & 586789 & 1976176 & 6765271 & 22624951 \\
	\hline
	\(380S(n-5) + 148 \) & \cellcolor{yellow} 638548 & 2008828 & \cellcolor{yellow} 6765288 & \cellcolor{yellow} 23160388 \\
	\hline
	\(1140S(n-6) + 528 \) & 611568 & 1915728 & 6026568 & 20295948 \\
	\hline
\end{tabular}
\end{center}
Except for 8, 9 and 13, the best lower bounds are obtained thanks to
the third inequality \( 380S(n-5) + 148 \leqslant S(n) \). The table
doesn't go any further, but the same inequality allows to improve the
lower bounds for every \( n \in [\![15,+\infty]\!] \).

\begin{corollary}
The growth rate for Schur (and Ramsey) numbers satisfies \(\gamma = \sqrt[5]{380} \geqslant 3.28 \).
\end{corollary}
\textsc{Proof :} It is a mere consequence of the inequality \(380S(n-5) + 148 \leqslant S(n) \). As for Ramsey's 
numbers growth rate, a lower bound can be found using Schur's one, thanks to \(S(n) \leqslant R_n(3)-2 \) 
(see \hyperlink{label1}{\textbf{[1]}}).

>>>>

\qquad In this section, we use Rowley's constructions [7] in a Schur context. To improve lower bounds for Ramsay's numbers, Rowley 
introduces partitions verifying some properties which can be extended using a method which generalizes Abbott-Hanson's [8] construction. 
Rowley named these partitions "templates", and we will keep this name in the entire article. We find suitable templates and use them
 to find new lower bounds for Schur numbers.

\subsection{Definition of S\(_+\)}

\begin{definition}
	We call SF-template of \(n\) colors and lenght \(p\) a partition of \( [\![1,p]\!]\) into \(n\) sum-free subsets \(A_1, A_2, ..., A_n\) which verify :
	\[
	\forall i \in [\![1, n-1]\!], \forall (x,y) \in A_i^2, x+y > p
	\Longrightarrow x+y-p \notin A_i
	\]
	We note \(S_+(n)\) the maximal lenght for a SF-template of \(n\) colors. 
\end{definition}

\begin{theorem}
	Let \(n \in [\![2, +\infty]\!]\), we have :
	\[
	2S(n-1)+1 \leqslant S_+(n) \leqslant S(n)
	\]
\end{theorem}

\begin{remark}
	\(S_+\) and \(S\) have the same asymptotic growth rate.
\end{remark}

\subsection{Inequalities on \(S_+\)}

The main result on \(S_+\) follows. It allows us to improve lower bounds on Schur numbers by computing \(S_+\).

\begin{theorem}
	Let $(n,k), (p,q) \in (\mathbb{N}^*)^2$. If there exists a \(SF\)-template of $k+1$ colors and lenght \(p\),
	and a partition of $n$ sum-free subsets of $[\![1,q]\!]$ then there exists a partition of $n+k$ sum-free subsets of $[\![1,pq]\!]$.
\end{theorem}

Setting $p = S_+(k+1)$ and $q = S(n)$ yields the following corollary.

\begin{corollary}
	Let \(n, k \in \mathbb{N}^*\), we have \\
	\[ S(n+k) \geqslant S_+(k+1)S(n) \]
\end{corollary}

Using a SAT solvers, we are able to exhibit SF-template,
hence providing lower bound on \(S_+\). Moreover, when using 
a SF-template in the above formula, it is possible to get an additive constant. We seek templates providing "affine"
inequalities : \(aS(n+k) + b \) where \(a\) is a lower bound for \(S_+(k+1)\). Further details can be found in the \hyperref[SAT]{SAT section}.
Here are the best inequalities on Schur numbers so far :
\[ S(n+1) \geqslant 3S(n) + 1 \]
\[ S(n+2) \geqslant 9S(n) + 4 \]
\[ S(n+3) \geqslant 33S(n) + 6 \]
\[ S(n+4) \geqslant 111S(n) + 43 \]
\[ S(n+5) \geqslant 380S(n) + 148 \]
\[ S(n+6) \geqslant 1140S(n) + 528 \]

The first inequality comes from the original Schur's paper \hyperlink{label1}{\textbf{[1]}}. The second one are due to Abott (référence nécessaire) 
and the third one to Rowley (reference nécessaire). The other ones are new. \\

We also have a similar theorem where only \(S_+\) is involved.

\begin{theorem}
	Let $(n,k), (p,q) \in (\mathbb{N}^*)^2$. If there exists a \(SF\)-template of $k+1$ colors and lenght \(p\),
	and \(SF\)-template of \(n\) color and lenght \(q\), then there exists \(SF\)-template of \((n+k)\) and lenght \(pq\).
\end{theorem}

@@

\begin{corollary}
	Let \(n, k \in \mathbb{N}^*\), we have \\
	\[ S_+(n+k) \geqslant S_+(k+1)S_+(n) \]
\end{corollary}

@@

The previous inequalities give new lower bounds for \(S(n)\) for
\( n \geqslant 9 \). We compute the lower
bounds for \( n \in [\![8,15]\!] \) using the four different inequalities, please notice that the best values for \( n = 8\) and \(n = 13\) were already obtained thanks to the first one, found by Rowley. The best lower bounds are highlighted.\\
\\
\begin{center}
\begin{tabular}{|*{5}{c|}}
    \hline
	\(n\) & 8 & 9 & 10 & 11 \\
	\hline
	\(33S(n-3) + 6 \) & \cellcolor{yellow} 5286 & 17694 & 55446 & 174444\\
	\hline
	\(111S(n-4) + 43 \) & 4927 & \cellcolor{yellow} 17803 & 59539 & 186523\\
	\hline
	\(380S(n-5) + 148 \) & 5088 & 16868 & \cellcolor{yellow} 60948 & \cellcolor{yellow} 203828 \\
	\hline
	\(1140S(n-6) + 528 \) & 5088 & 15348 & 50688 & 182928\\
	\hline
	\hline
	\(n\) & 12 & 13 & 14 & 15 \\
	\hline
	\(33S(n-3) + 6 \) & 587505 & \cellcolor{yellow} 2011290 & 6726330 & 21072090\\
	\hline
	\(111S(n-4) + 43 \) & 586789 & 1976176 & 6765271 & 22624951 \\
	\hline
	\(380S(n-5) + 148 \) & \cellcolor{yellow} 638548 & 2008828 & \cellcolor{yellow} 6765288 & \cellcolor{yellow} 23160388 \\
	\hline
	\(1140S(n-6) + 528 \) & 611568 & 1915728 & 6026568 & 20295948 \\
	\hline
\end{tabular}
\end{center}
Except for 8, 9 and 13, the best lower bounds are obtained thanks to
the third inequality \( S(n+5) \geqslant 380S(n) + 148\). The table
doesn't go any further, but the same inequality allows to improve the
lower bounds for every \( n \geqslant 15 \).

\begin{corollary}
The growth rate for Schur (and Ramsey) numbers satisfies \(\gamma \geqslant \sqrt[5]{380} \approx 3.28 \).
\end{corollary}
\textsc{Proof :} It is a mere consequence of the inequality \( S(n+5) \geqslant 380S(n) + 148\). As for Ramsey's 
numbers growth rate, a lower bound can be found using Schur's one, thanks to \(S(n) \leqslant R_n(3)-2 \) 
(see \hyperlink{label1}{\textbf{[1]}}).
